---
title: "rmats-coronary-analysis"
author: "Shay Iyer"
date: "2025-07-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

# This was just copy-pasted as needed

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "show")
library(tidyverse)
library(maser)
library(rtracklayer)
```

Group 1: Ischemic
Group 2: Non-ischemic

# Maser

```{r}
ca.maser <- maser("/Volumes/cphg-millerlab/katherine_shillito/coronary-rmats/03_outputs/post/post-directory1", c("Ischemic", "Non-ischemic"), ftype="JCEC")
```

## Maser data pre-processing

```{r}
ca.filt <- filterByCoverage(ca.maser, avg_reads=5) 
ca.top <- topEvents(ca.filt, fdr = 0.1, deltaPSI = 0.05)
ca.filt.df <- summary(ca.filt)
ca.maser.df <- summary(ca.maser)

ca.top.df <- summary(ca.top) %>% arrange(desc(abs(IncLevelDifference)))

ca.filt.df <- ca.filt.df %>% arrange(desc(abs(IncLevelDifference)))
```

## Basic plots

```{r}
se <- summary(ca.top, type = "SE") %>% mutate("eventType" = "SE") %>% select(c(eventType, ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand))
mxe <- summary(ca.top, type = "MXE") %>% mutate("eventType" = "MXE") %>% select(c(eventType, ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand))
a3ss <- summary(ca.top, type = "A3SS") %>% mutate("eventType" = "A3SS") %>% select(c(eventType, ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand))
a5ss <- summary(ca.top, type = "A5SS") %>% mutate("eventType" = "A5SS") %>% select(c(eventType, ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand))
ri <- summary(ca.top, type = "RI") %>% mutate("eventType" = "RI") %>% select(c(eventType, ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand))

allevents <- rbind(se, mxe, a3ss, a5ss, ri)

allevents.filt <- allevents %>% filter(FDR < 0.05, abs(IncLevelDifference) > 0.1)
```

```{r}
allevents <- allevents %>%
  mutate(Significant = ifelse(PValue < 0.05 & FDR < 0.05 & abs(IncLevelDifference) > 0.2, "Significant", "Not Significant")) %>% mutate(SigDirection = case_when(
    PValue < 0.05 & FDR < 0.05 & IncLevelDifference > 0.2  ~ "Ischemic",
    PValue < 0.05 & FDR < 0.05 & IncLevelDifference < -0.2 ~ "Non-ischemic",
    TRUE ~ "Not Significant"
  ))

ggplot(allevents.filt, aes(x = eventType)) +
  geom_bar() +
  labs(title = "Breakdown of significant splicing events by event type",
       subtitle = "Ischemic vs. non-ischemic samples; FDR < 0.05, |∆PSI|>0.1", 
       x = "Event Type",
       y = "Number of Events") +
  theme_minimal()
```


```{r}
top_hits <- allevents %>%
  filter(SigDirection != "Not Significant") %>%
  arrange(PValue) %>%
  head(25)

library(ggrepel)
ggplot(allevents, aes(x = IncLevelDifference, y = -log10(PValue))) +
  geom_point(aes(color = SigDirection), alpha = 0.7) +
  scale_color_manual(values = c("Ischemic" = "red", "Non-ischemic" = "blue", "Not Significant" = "grey")) +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  # Add gene labels for significant points
  geom_text_repel(
    data = top_hits,
    aes(label = geneSymbol),
    size = 3,
    max.overlaps = 1000,      # Increase if some labels still missing
    box.padding = 0.35,       # Adjust space around labels
    point.padding = 0.3,      # Adjust space between label and point
    force = 2,                # Increase for stronger repulsion
    segment.size = 0.2        # Thinner connector lines
  ) +
  labs(title = "Volcano Plot of Top 25 most significant rMATS Events",
       subtitle = "FDR < 0.05, p < 0.05, |∆PSI| > 0.18; arranged by p value",
       x = expression(Delta*PSI),
       y = expression(-log[10](PValue)),
       color = "Condition") +
  theme_minimal()

```

## Specific splicing events


```{r}

se.filt.sum <- summary(ca.filt, type = "SE") 
mxe.filt.sum <- summary(ca.filt, type = "MXE")
a3ss.filt.sum <- summary(ca.filt, type = "A3SS")
a5ss.filt.sum <- summary(ca.filt, type = "A5SS")
ri.filt.sum <- summary(ca.filt, type = "RI")
se.top.sum <- summary(ca.top, type = "SE") %>% arrange(IncLevelDifference) %>% select(c(ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand)) %>% mutate(eventType = "SE")
mxe.top.sum <-summary(ca.top, type = "MXE") %>% arrange(IncLevelDifference) %>% select(c(ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand)) %>% mutate(eventType = "MXE")
a3ss.top.sum <- summary(ca.top, type = "A3SS") %>% arrange(IncLevelDifference) %>% select(c(ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand)) %>% mutate(eventType = "A3SS")
a5ss.top.sum <- summary(ca.top, type = "A5SS") %>% arrange(IncLevelDifference) %>% select(c(ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand)) %>% mutate(eventType = "A5SS")
ri.top.sum <- summary(ca.top, type = "RI") %>% arrange(IncLevelDifference) %>% select(c(ID, GeneID, geneSymbol, PValue, FDR, IncLevelDifference, PSI_1, PSI_2, Chr, Strand)) %>% mutate(eventType = "RI")

all.top.events <- rbind(se.top.sum, mxe.top.sum, a3ss.top.sum, a5ss.top.sum, ri.top.sum)

```

> Genes from top events implicated in VSMC transitions/atherosclerosis, as identified by CellGPT 4o

- FHOD1: [Novel signaling axis of FHOD1-RNF213-Col1α/Col3α in the pathogenesis of hypertension-induced tunica media thickening](https://www.sciencedirect.com/science/article/pii/S0022282823001141), [Formin homology domain-containing protein 1 regulates smooth muscle cell phenotype](https://pubmed.ncbi.nlm.nih.gov/21106951/)
- STAT6: ["The downstream network of STAT6 in promoting vascular smooth muscle cell phenotypic switch and neointimal formation"](https://onlinelibrary.wiley.com/doi/full/10.1002/cbin.12056)
- MRTFA: [Signaling Mechanisms That Regulate Smooth Muscle Cell Differentiation](https://pmc.ncbi.nlm.nih.gov/articles/PMC3141215/) (this also mentions FHOD1), [A Novel Role of Brg1 in the Regulation of SRF/MRTFA-dependent Smooth Muscle-specific Gene Expression](https://www.sciencedirect.com/science/article/pii/S0021925820746466?via%3Dihub)
- SRGAP1: [H3K4 di-methylation governs smooth muscle lineage identity and promotes vascular homeostasis by restraining plasticity (Fig 8c)](https://pmc.ncbi.nlm.nih.gov/articles/PMC8567421/)
- IL32: ["Clinical Implications of IL-32, IL-34 and IL-37 in Atherosclerosis: Speculative Role in Cardiovascular Manifestations of COVID-19"](https://pmc.ncbi.nlm.nih.gov/articles/PMC8377289/?utm_source=chatgpt.com)
- DDIT3 (aka CHOP): [CHOP Increases TRIB3-Dependent miR-208 Expression to Potentiate Vascular Smooth Muscle Cell Proliferation and Migration by Downregulating TIMP3 in Atherosclerosis](https://pubmed.ncbi.nlm.nih.gov/33856595/)


```{r}
# Specific splicing events
print("------FHOD1------")
ca.fhod1 <- geneEvents(ca.top, geneS = "FHOD1", fdr = 0.05, deltaPSI = 0.1) # 1 SE 
```

```{r}
fhod1.violin <- plotGenePSI(ca.fhod1, type = "SE", show_replicates = TRUE) + labs(
  title = "PSI values for significant SE event, FHOD1",
  subtitle = "FDR < 0.05, p < 0.05, |∆PSI| > 0.1")
```

```{r}

```
